---
title: "Home"
output: html_document
runtime: shiny
resource_files:
- _site.yml
- footer.html
---


```{r setup, include=FALSE}
library(shiny)
library(shinythemes)
library(data.table)
library(tidyverse)
library(rvest)
library(shiny)
library(plotly)
library(chron)

df <- data.table::fread("https://media.githubusercontent.com/media/flemm0/PM566_project/main/data/processed_data.csv")
```

--------------

### Introduction

In this project, I analyzed dietary interview [data](https://www.ars.usda.gov/northeast-area/beltsville-md-bhnrc/beltsville-human-nutrition-research-center/food-surveys-research-group/docs/wweia-documentation-and-data-sets/) from the National Health and Nutrition Examination Survey, also known as NHANES. The goal for this analysis was to investigate the factors that affect peoples' consumption of high amounts of sugar and saturated fatty acids. According to articles published by the CDC on diabetes, people should be wary of continued consumption of foods containing high amounts of these ingredients, as they contribute to the development of insulin resistance and prediabetes. 


### About the data

Data for the NHANES program is collected on an annual cycle. However, the program suspended operations in March 2020 due to the pandemic. The data I used for this analysis was a combination of the NHANES 2017-2018 cycle and 2019 to March 2020.


The data I used came mainly from two files, storing participants' answers to day 1 and day 2 of the questionnaire. The rows in each files are separated as one food/drink item. The other columns describe information such as which participant ate/drank this item, when it was consumed, estimates on number of calories and macronutrients, and much more information. 

I used two other files to get demographic information, and to link food codes to actual descriptions of the food/drink items in the questionnaire answer file. 


### Questions

1. **What time and/or day of the week do people generally eat foods high in sugar or saturated fatty acids (fa)?**
2. **Does sugar or saturated fa consumption vary by age, ethnicity, or gender?**
3. **Does the source of food or whether the meal was eaten at home have an effect on sugar or saturated fa consumption?**
4. **Finally, what specific food items are associated with high amounts of sugar or saturated fa?**


--------------

This website provides interactive visualizations meant to supplement the full report. If you are interested in reading more, click [here](https://github.com/flemm0/nhanes_survey_analysis/raw/main/full_report/full_report.pdf)

If you are interested in viewing the source code for the website and report, click [here](https://github.com/flemm0/nhanes_survey_analysis)

--------------


```{css style, echo=FALSE}
.column-left{
  float: left;
  width: 15%;
  text-align: left;
  padding: 5px;
}
.column-right{
  float: right;
  width: 85%;
  text-align: left;
  padding: 5px;
}
```


{.column-left}
--------------

```{r sidebar, echo=FALSE}
selectizeInput(
              inputId = "ingredient", 
              label = "Select Data to View:", 
              choices = c("Sugar", "Saturated Fatty Acid"),
              selected = "Sugar"
              )

```


{.column-right}
--------------

##### Figure 1

```{r p1, echo=FALSE}

renderPlotly({
      x = ifelse(input$ingredient == "Sugar", 
               "total_sugars_gm", "total_saturated_fatty_acids_gm")
    
    df$eating_time <- as.POSIXct(df$time_of_eating_occasion_hhmm, format = "%H:%M:%S")
    df %>%
      select(intake_day_of_the_week, intake_day_cat, time_of_eating_occasion_hhmm, eating_time, !! rlang::sym(x))   %>%
      mutate(eating_time = format(round(eating_time, units="hours"), format="%H:%M")) %>%
      group_by(intake_day_cat, eating_time) %>%
      summarize(m = mean(!! rlang::sym(x))) %>%
      arrange(match(intake_day_cat, 
                    c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")),
              eating_time) %>%
      plot_ly(x = ~eating_time, y = ~m, 
              hoverinfo = "text", 
              text = ~paste( 
                paste(intake_day_cat, ":", sep = ""),
                paste("Eating time:", eating_time, "(rounded to nearest hour)", sep= " "),
                paste("Average", tolower(input$ingredient), "consumption (grams):", round(m, digits = 3), sep =
                        " "),
                sep = "<br>"
      )) %>%
      add_lines(color = ~intake_day_cat) %>%
      layout(
        title = paste0("Average ", input$ingredient, " Consumption By Time of Day"),
        yaxis = list(title = "Average Consumption (grams)"),
        xaxis = list(
          title = "Eating Time",
          type = "time",
          tickformat="%H:%M:%S"
          )
        )
  })
```

The figure above plots a rolling average of how sugar or saturated fatty acid consumption change by the time of day and are grouped by the day of the week. Sugar consumption starts off lower at midnight on Tuesday and Thursday, and are higher for the other days of the week. Then, the lines seem to converge at around 7:00 AM, and from there there are spikes in sugar consumption at around 10:00 AM, 3:30 PM, and 9:00 - 11:00 PM. Saturated fatty acid consumption varies less by day of the week with the lines converging at around 5:00 AM and steadily increasing for the next 21 hours.

--------------

##### Figure 2

```{r p2, echo=FALSE}

renderPlotly({
    x = ifelse(input$ingredient == "Sugar", 
              "total_sugars_gm", "total_saturated_fatty_acids_gm")
    lvls <- df[, eating_occasion := 
              fifelse(eating_occasion == "Desayano", "Desayano (breakfast)",
              fifelse(eating_occasion == "Almuerzo", "Almuerzo (breakfast)",
              fifelse(eating_occasion == "Comida", "Comida (lunch)",
              fifelse(eating_occasion == "Merienda", "Merienda (snack)",
              fifelse(eating_occasion == "Cena", "Cena (dinner)",
              fifelse(eating_occasion == "Entre comida", "Entre comida (snack)",
              fifelse(eating_occasion == "Botana", "Botana (snack)",
              fifelse(eating_occasion == "Bocadillo", "Bocadillo (snack)",
              fifelse(eating_occasion == "Tentempie", "Tentempie (snack)",
              fifelse(eating_occasion == "Bebida", "Bebida (drink)", eating_occasion))))))))))] %>%
  group_by(eating_occasion) %>%
  summarise(m = median(!! rlang::sym(x))) %>%
  arrange(m) %>%
  pull(eating_occasion)
    
  df[, eating_cat := 
     fifelse(grepl("snack", eating_occasion, ignore.case = TRUE), "snack", 
             fifelse(grepl("(breakfast|lunch|dinner|supper|brunch)", eating_occasion, ignore.case = TRUE), 
                     "meal", "other"))] %>%
    mutate(consumption = sqrt(!!rlang::sym(x))) %>%
    select(eating_cat, eating_occasion, consumption) %>%
    plot_ly(x = ~consumption, y = ~factor(eating_occasion, lvls), color = ~eating_cat) %>%
    add_boxplot() %>%
    layout(
      title = paste("Distribution of", input$ingredient, "Consumption By Eating Occasion", sep = " "),
      yaxis = list(title = "Eating Occasion"),
      xaxis = list(title = paste("Square Root", input$ingredient, "Consumption", sep = " "), hoverformat = ".2f")
    )
  })
```

The figure above plots the distribution of sugar or saturated fatty acid consumption grouped by what eating occasion. The x-axis has been square root transformed, to account for the highly skewed consumption observations. Although eating occasions at different times of the day, such as breakfast, lunch, and dinner do not see much difference in sugar or saturated fatty acid consumption, the plots above do reveal that average sugar consumption is lowest in formal meals such as breakfast, lunch or dinner. On the other end, snacking occasions typically involve much higher average sugar consumption. The reverse trend can be seen for average saturated fatty acid consumption, as snacks have lower medians than meals such as lunch or dinner.

--------------

##### Figure 3

```{r p3, echo=FALSE}
renderPlotly({
    x = ifelse(input$ingredient == "Sugar", 
              "total_sugars_gm", "total_saturated_fatty_acids_gm")
    
    df %>%
      select(racehispanic_origin_w_nh_asian,
             age_category, 
             gender, 
             respondent_sequence_number, 
             interview_day,
             !!rlang::sym(x)) %>%
      group_by(racehispanic_origin_w_nh_asian, age_category, gender) %>%
      summarize(total_consumption = sum(!!rlang::sym(x)), count = n()) %>%
      mutate(tc_scaled = total_consumption / count) %>%
      group_by(age_category, gender) %>%
      mutate(prop = tc_scaled / sum(tc_scaled)) %>%
      ungroup() %>%
      unite(age_gender, c(age_category, gender), sep = " ") %>%
      plot_ly(
        y = ~prop, 
        x = ~factor(age_gender, levels = c(
                         "<1 male", "<1 female",
                         "1-3 male", "1-3 female",
                         "4-8 male", "4-8 female",
                         "9-13 male", "9-13 female", 
                         "14-18 male", "14-18 female",
                         "19-30 male", "19-30 female",
                         "31-50 male", "31-50 female",
                         "51-70 male", "51-70 female",
                         "70+ male", "70+ female")), 
        color = ~racehispanic_origin_w_nh_asian, 
        type = "bar",
        hoverinfo = "text",
        text = ~paste(
          paste("Age category (years):", age_gender, sep = " "),
          paste("Proportion of", tolower(input$ingredient), "consumption:", round(prop*100, digits = 1), "%"),
          sep = "<br>"
        )) %>% 
      layout(barmode = "stack", 
             xaxis = list(title = "Age Category (in years) and Gender"), 
             yaxis = list(title = paste("Proportion of", input$ingredient, "Consumption", sep = " ")),
             title = paste("Proportion of", input$ingredient, "Consumption by Age, Gender, and Ethnicity", sep = " "))
  })
```

The figure above plots the proportion of sugar or saturated fatty acid consumption within each ethnicity category, separated by age and gender. The proportion of consumption appears to be relatively similar across all ethnicity groups, with each group having about 17% of the consumption. The non-hispanic asian group seems to be slightly less than others, however, with an average of about 13% of the proportion. 

--------------

##### Figure 4

```{r echo=FALSE}
renderPlotly({
    x = ifelse(input$ingredient == "Sugar", 
              "total_sugars_gm", "total_saturated_fatty_acids_gm")
    
    lvls <- df %>%
      filter(!is.na(food_source) & !is.na(!! rlang::sym(x))) %>%
      group_by(food_source) %>%
      summarise(m = median(sqrt(!! rlang::sym(x)))) %>%
      arrange(m) %>%
      pull(food_source)
    
    df %>%
      filter(!is.na(food_source) & !is.na(!! rlang::sym(x))) %>%
      mutate(consumption = round(sqrt(!! rlang::sym(x)), 2)) %>%
      plot_ly(x = ~consumption, y = ~factor(food_source, lvls)) %>%
      add_boxplot() %>%
      layout(
        yaxis = list(title = "Source of Food", tickfont = list(size = 8)),
        xaxis = list(title = paste("Square Root of", input$ingredient, "Consumption", sep = " ")),
        title = paste("Distribution of", input$ingredient, "Consumption By Food Source", sep = " ")
      )
  })
```

It is interesting to note that the median for both sugar and saturated fatty acid distributions is highest from food obtained from fundraiser sales. School cafeterias also are prominent sources of high sugar and saturated fatty acid consumption. Additionally, the highest consumption of sugar and saturated fatty acid consumption comes from the "Store - grocery/supermarket" category.

